\section{Introduction}
\label{sec:intro}

The integration of Large Language Models (LLMs) into software engineering has progressed rapidly through three distinct phases: suggestion (e.g., inline autocompletion), instruction (e.g., conversational chat interfaces), and delegation (e.g., autonomous agents). In this current phase of delegated execution, an agent is not merely a generator of code snippets; it is an active developer. Equipped with a tool belt of shell executors, file readers, linters, compilers, and test suites, the agent operates directly on target repositories, running iterative loops to edit files, compile binaries, parse build logs, and debug errors.

However, as agentic workflows evolve from simple scripts to long-running, concurrent, multi-agent systems, their durable coordination mechanisms remain underdeveloped. Many contemporary workflows still depend heavily on the natural-language prompt, one workbench conversation, and the final git diff. A session or tool switch therefore changes not only the executing process but often the available project memory, ownership state, and evidence.

We argue that this transcript-only paradigm is fundamentally inadequate for serious software engineering environments. Chat transcripts are verbose, non-deterministic, and prone to context window drift. More critically, they lack clear boundaries of authority, fail to capture environment validation results in a machine-checkable format, and provide no guarantees that code changes satisfy the intended constraints. Once an agent is granted write access to a repository and the authority to run arbitrary CLI tools, software engineering transforms from a language-modeling problem into a \textbf{distributed execution and security problem}.

When viewed through a systems-oriented lens, agentic software engineering shares core challenges with classic distributed systems:
\begin{itemize}
    \item \textbf{Authority \& Isolation}: How do we ensure that an agent only mutates the directories it is authorized to touch, without leaking credentials or accessing sensitive environment variables?
    \item \textbf{Concurrency}: How do we reduce duplicate task starts, isolate mutations, respect dependencies, and detect integration failure without pretending worktrees guarantee compatible merges?
    \item \textbf{Failure, Handoff \& Recovery}: If an agent runs out of tokens, changes workbenches, or crashes mid-task, how can a subsequent process recover intent, unresolved state, ownership, and proof without replaying one vendor conversation?
    \item \textbf{Auditability}: How can a human developer efficiently verify that an agent's changes are correct without manually reviewing thousands of lines of conversational LLM reasoning?
\end{itemize}

To address these challenges, this paper formalizes four core primitives for accountable execution:
\begin{enumerate}
    \item \textbf{Intent}: Durable task objectives, constraints, open questions, and proof expectations represented outside the model's transient context.
    \item \textbf{Custody}: Exclusive task ownership and isolated git-worktree or container boundaries within which the agent operates.
    \item \textbf{Trajectory}: Event-backed records of task, broker, proof, and governance activity that can be rendered as a flight-recorder timeline.
    \item \textbf{Proof}: Programmatically checkable validation, proof-run, workunit, and provenance evidence; cryptographic signing is a future extension, not an implemented claim here.
\end{enumerate}

We present \textbf{Decapod}, a daemonless, local-first governance kernel, as a reference implementation of this model. The primary experiment remains one-shot walk-away delegation: whether, under the same complex natural request, pre-inference resolution of authoritative context lets an unchanged model converge toward the intended outcome with fewer failed cycles and less human intervention. We also develop a second systems thesis: Intent, Custody, Trajectory, and Proof may produce \emph{fleet coherence}, allowing independently launched agent processes to share durable project authority, transfer work, avoid duplicated cognition, and publish against common evidence without sharing one conversation or long-running orchestrator. Agents and conversations are ephemeral execution processes; the governed repository is the durable coordination subject.

The implementation mapping is pinned to Decapod \texttt{e7df80d} (\texttt{v0.66.1}). Prospective Studies B--D test handoff, concurrent work, and tool switching; Study E is a non-causal longitudinal case study. The contribution remains pre-results: mechanisms, boundaries, schemas, and falsifiable protocols rather than measured treatment effects.
