\section{Formal Execution Model}
\label{sec:model}

To resolve the five failure modes identified in Section~\ref{sec:problem}, we formalize an execution model built around Intent, Custody, Trajectory, and Proof. This model abstracts agentic software engineering as a state transition system operating over a repository space.

Let $R$ represent the state of the repository, including the files, directory structure, git history, and runtime environment. An agent run is a sequence of state transitions $R_0 \xrightarrow{a_1} R_1 \xrightarrow{a_2} \dots \xrightarrow{a_n} R_n$, where each action $a_i$ is executed by the agent via tools.

\subsection{Intent: Desired Outcome and Resolution}
Intent ($I$) is the human or team's desired outcome, why it matters, the constraints and priorities that must remain true, unresolved questions, stop conditions, and standard of completion:
$$I = \langle O, M, C, Q, S, A \rangle$$
where $O$ is the desired outcome, $M$ its motivation, $C$ constraints and priorities, $Q$ unresolved questions, $S$ stop conditions, and $A$ acceptable completion. A prompt is an initial and often incomplete expression of $I$. Plans, todos, specifications, context capsules, proof hooks, and rubrics are governed projections or resolutions of intent, not its source of authority. Decapod distributes these representations across repository artifacts so intent can be re-resolved against repository state and human escalation points without requiring the initial prompt to contain a complete procedural specification.

Before implementation inference, the governed system resolves a context set $X = \langle U, A_e, E, D, Q_e, V \rangle$, where $U$ contains selected authority and constitution nodes, $A_e$ contains architecture and project specifications, $E$ contains capabilities and constraints, $D$ contains prior decisions and applicable standards, $Q_e$ contains unresolved questions and escalation rules, and $V$ contains validation and proof expectations. Each supplied item has an authority source, selection rationale, content hash, and relevance record. The primary experiment tests this context-alignment mechanism against a conventional run receiving the identical natural prompt, not a different model.

\subsection{Custody: Environment Bounding}
Custody ($K$) defines the execution and security sandbox for the agent. Formally, it is a tuple:
$$K = \langle W, B, A, P \rangle$$
where:
\begin{itemize}
    \item $W \subset R$ is the isolated workspace slice (a task-scoped git worktree and, when configured, a container).
    \item $B$ is the task-specific branch locked for the agent.
    \item $A$ is the set of governance operations and proof capabilities available to the run.
    \item $P$ represents the repository and runtime boundary enforced by the workspace configuration.
\end{itemize}
By enforcing custody boundaries, the kernel keeps the agent's repository mutations in $W$ and prevents direct work on protected branches. This reduces cross-task workspace collisions; it is not a complete operating-system sandbox or credential-isolation proof.

\subsection{Trajectory: The Governance Record}
The Trajectory ($T$) is an event-backed sequence of governance records:
$$T = [t_1, t_2, \dots, t_n]$$
where a record may include timestamp, operation, actor, session, correlation identifier, status, and structured details. Decapod journals todo, broker, proof, and related governance events and can render them through its flight recorder. These records improve reconstruction and review, but they are not asserted here to be a complete capture of every model token, shell byte, file diff, or tamper-proof execution event.

\subsection{Proof: Completion Evidence}
A Proof ($P_f$) is the set of machine-checkable evidence attached to a governed task:
$$P_f = \langle I, R_n, \{\operatorname{Pass}(v_j)\}, E \rangle$$
where $E$ contains relevant proof events, workunit results, validation output, and provenance references. In Decapod, validation and proof gates can block promotion or completion paths when required evidence is absent. A proof record establishes what the configured gates observed; it does not establish semantic correctness beyond those gates, and this paper makes no claim that the current implementation signs a universal certificate.

\subsection{Heterogeneous Multi-Agent Layer}

The single-run transition sequence is insufficient for independently launched processes. Let $\mathcal{A}=\{\alpha_1,\ldots,\alpha_m\}$ be agent processes, $\mathcal{M}$ the underlying models, $\mathcal{W}$ the workbenches or runtimes, and $G=(V,E_G)$ a governed task/dependency graph. Each agent has a model/workbench pair, but neither is the durable project subject. We define:
\begin{align*}
o &: V \rightarrow \mathcal{A}\cup\{\varnothing\} && \text{primary task ownership},\\
w &: V \rightarrow \mathcal{S} && \text{task-to-workspace assignment},\\
C(v) &&& \text{resolved authoritative context},\\
T(v) &&& \text{accumulated governance trajectory},\\
P(v) &&& \text{proof and validation state},\\
H(v,\alpha_i,\alpha_j) &&& \text{governed ownership transfer},\\
Q &&& \text{integration and publication state}.
\end{align*}
The repository state includes the durable representations of $G,o,w,C,T,P,$ and $Q$; a workbench conversation does not. A handoff $H$ changes ownership and records durable transfer information while the receiving process resolves the current repository state. It cannot transfer inaccessible hidden reasoning.

\subsection{Qualified Fleet Invariants}

The implementation audit distinguishes enforced, conditional, and desired properties:
\begin{enumerate}
    \item \textbf{Exclusive primary custody (enforced for exclusive mode):} an atomic exclusive claim conflicts with another active primary assignee. Trust-gated shared secondary ownership exists, so uniqueness is not unconditional.
    \item \textbf{Workspace custody (enforced on governed paths):} a todo-scoped worktree is required and protected/root checkout work is blocked. External tools can bypass Decapod, and containers are configuration-dependent.
    \item \textbf{Dependency awareness (represented and partially gated):} dependencies are durable and replayable; this is not a globally serializable scheduler.
    \item \textbf{Handoff continuity (implemented but bounded):} approved ownership transfer preserves task assignment, summary/event, durable knowledge, and available repository references. Complete conversation state, unresolved-state completeness, and merge success are not guaranteed.
    \item \textbf{Provider-independent project state (implemented as an external contract):} generated entrypoints route multiple workbenches to one repository contract. Compliance by every workbench is empirical, not enforced by Decapod.
    \item \textbf{Proof-gated publication (enforced on the Decapod publish path):} provenance manifests and configured workunit/evaluation gates are required. Task-local evidence remains insufficient if post-integration gates fail or are absent.
    \item \textbf{Projection consistency (enforced for declared surfaces):} validation detects known derived-state drift. This is not global distributed consensus.
\end{enumerate}

We define \emph{fleet coherence} as the degree to which these bounded properties let heterogeneous processes preserve common authority, avoid conflicting custody, reconstruct relevant state, and share an acceptance boundary. It is an emergent property of Intent, Custody, Trajectory, and Proof, not a fifth primitive.
