\section{Decapod: A Reference Governance Kernel}
\label{sec:system}

Decapod is a daemonless, local-first governance kernel for AI coding agents. It is invoked by agents at governance boundaries; it does not replace the model's tool loop or alter model weights. The implementation mapping is pinned to commit \texttt{207e9a3} (tag \texttt{v0.66.3}); the companion claim matrix lists command, source, test, status, boundary, and planned measure for each research concept. The changes from the previously inspected \texttt{v0.66.1} revision through this release are release, CI, and research-link documentation changes; the mapped implementation and test paths are unchanged.

\subsection{System Architecture}
The CLI and Decapod-specific structured process-RPC surfaces route requests into core subsystems that persist state under \texttt{.decapod/}. The principal surfaces relevant to this paper are:
\begin{enumerate}
    \item \textbf{Intent and coordination}: governed plans, todos, specifications, context capsules, knowledge, and proof hooks progressively resolve human intent. The todo subsystem records claims, tags/categories, dependencies, presence, owners, and approved handoffs; none of these projections replaces human or team authority.
    \item \textbf{Custody}: workspace management creates todo-scoped git worktrees and, when configured, container workspaces. Protected branches and workspace checks prevent an agent from treating the human root checkout as its execution surface.
    \item \textbf{Trajectory}: todo, broker, proof, federation-memory, and related event journals are rendered by the flight recorder as a timeline or transcript. This is a selected governance record, not a complete capture of all model or shell activity.
    \item \textbf{Proof and validation}: configurable proof commands produce structured results and proof events; \texttt{decapod validate}, workunit gates, provenance artifacts, and verification checks supply promotion evidence.
\end{enumerate}

The generated specs describe the same corridor as intent $\rightarrow$ claim $\rightarrow$ workspace $\rightarrow$ proofs $\rightarrow$ publish. The public crate's cloud backend explicitly reports unavailable; cloud-mode initialization writes only an experimental registration payload while local state remains usable. Remote leases, synchronization, and cloud handoff are therefore future work, not evaluated capabilities.

\subsection{Intent and Governed Workflow}
Decapod does not currently require a universal \texttt{.decapod/intent.json} manifest or file-marker completion protocol. Instead, the workflow is distributed across explicit interfaces such as:
\begin{verbatim}
decapod session acquire
decapod todo add "task description"
decapod todo claim --id <task-id>
decapod todo handoff --id <task-id> --to <agent> --summary "..."
decapod infer orientation --intent "..." --task-id <task-id>
decapod workspace ensure --container
decapod validate
decapod todo done --id <task-id> --validated
\end{verbatim}
Governed plans add a stronger pre-execution contract: execution is blocked when the plan is not approved, when intent or unknowns remain unresolved, or when no todo is selected. These mechanisms operationalize durable intent without implying that every natural-language requirement is machine-verifiable.

\subsection{Constitutional Authority and Context-Poisoning Boundary}
The baseline constitution is compiled into the Decapod binary and exposed as structured, queryable nodes. A binding project-local override is merged with tested project precedence. Context resolution then produces task-scoped, provenance-bearing projections of this authority. Thus the constitution is not merely another repository instruction file: the intended chain distinguishes embedded baseline authority, explicit override, and selected context supplied before implementation inference. Exact asset, command, and test paths appear in the companion implementation matrix.

This is Decapod's proposed answer to repository-instruction poisoning: authority, scope, negative requirements, and proof expectations should be queryable and auditable rather than treating every available instruction as equally trustworthy. It does not prove that supplied context is relevant or benign, and project-controlled surfaces can still be hostile. The current code has no organization-level overlay; organization policy, three-level precedence, revocation, and conflict semantics remain roadmap work.

\subsection{Operator Decoupling}
Generated provider/workbench entrypoint shims delegate to one universal \texttt{AGENTS.md} contract. The intended human interface remains natural language: the operator states the outcome to the available agent, and the agent invokes Decapod's CLI or structured RPC at governance boundaries. This is interaction-contract portability, not a claim that vendor-specific commands disappear or that every workbench obeys the contract.

\subsection{Candidate Agent-Governance Interoperability Profile}
We offer the governance boundary as a candidate thin-waist profile: versioned semantics for capability discovery, constitutional authority, task custody and handoff, workspace status, selected trajectory, proof, and publication. Coding-agent harnesses could pre-bundle or discover a conforming local component so a user and project can move among Claude, Codex, Antigravity, Grok, and other workbenches without making one vendor conversation authoritative.

The implemented base is narrower. \texttt{decapod rpc} accepts a Decapod-specific structured request/response document over process standard input/output. It is not JSON-RPC 2.0, an MCP server, an A2A endpoint, or an HTTP service; no vendor-bundling agreement is claimed. MCP supplies a host--tool lifecycle and transport model~\cite{mcp2025architecture}. A2A supplies agent discovery and task/artifact exchange~\cite{a2a2026spec}. HTTP and IP illustrate narrow layered interfaces that permit independent evolution~\cite{rfc9110http,rfc791ip}, but the analogy does not imply comparable maturity, governance, or adoption. Native adapters, negotiation, conformance vectors, identity, and version-skew behavior are prospective.

As of a July 13, 2026 API snapshot, Decapod had 222 GitHub stars, 21 forks, and 6,983 cumulative crates.io downloads~\cite{github2026decapod,cratesio2026decapod}. These counts establish public distribution and interest only, not correct use, interoperability, or empirical efficacy.

The same authenticated audit verified Decapod governance evidence in 16 repositories across three owner namespaces (13 public and three private). Fifteen expose \texttt{.decapod/config.toml}; the public \href{https://github.com/alexhraber/alexhraber}{\nolinkurl{alexhraber/alexhraber}} profile repository exposes a generated Decapod \texttt{AGENTS.md}:
\begin{sloppypar}
\begin{itemize}
    \item \textbf{alexhraber:} \href{https://github.com/alexhraber/alexhraber}{\nolinkurl{alexhraber}}, \href{https://github.com/alexhraber/pi-cluster}{\nolinkurl{pi-cluster}}, \href{https://github.com/alexhraber/builddeck}{\nolinkurl{builddeck}}, \href{https://github.com/alexhraber/agentic-sdlc-intake-attack}{\nolinkurl{agentic-sdlc-intake-attack}},\\
    \href{https://github.com/alexhraber/metaplay-fake-interview-supply-chain-incident}{\nolinkurl{metaplay-fake-interview-supply-chain-incident}}, \href{https://github.com/alexhraber/go-scaling}{\nolinkurl{go-scaling}}, \href{https://github.com/alexhraber/nixos}{\nolinkurl{nixos}}, private \href{https://github.com/alexhraber/cycle}{\nolinkurl{cycle}}, and private \href{https://github.com/alexhraber/mentora-automa}{\nolinkurl{mentora-automa}};
    \item \textbf{worldofgeese:} \href{https://github.com/worldofgeese/den}{\nolinkurl{den}}, \href{https://github.com/worldofgeese/tarot-api}{\nolinkurl{tarot-api}}, and \href{https://github.com/worldofgeese/mythras-chargen}{\nolinkurl{mythras-chargen}};
    \item \textbf{DecapodLabs:} \href{https://github.com/DecapodLabs/decapod}{\nolinkurl{decapod}}, private \href{https://github.com/DecapodLabs/gensor}{\nolinkurl{gensor}}, \href{https://github.com/DecapodLabs/accountable-agentic-execution}{\nolinkurl{accountable-agentic-execution}}, and \href{https://github.com/DecapodLabs/pincher}{\nolinkurl{pincher}}.
\end{itemize}
\end{sloppypar}
The profile README is the author's public researcher/engineer credential surface; that is a professional provenance claim, not cryptographic identity authentication. Because these repositories are maintainer-correlated, the inventory establishes deployment footprint rather than 16 independent adopters, active protocol compliance, or a treatment effect. The companion inventory records its verification boundary.

The author reports a consistent lifecycle across new projects and engineering efforts: run \texttt{decapod init} once when establishing a repository, before the first agent interaction or other engineering operation. Thereafter the human interfaces naturally with one or more agents, while the generated \texttt{AGENTS.md} contract requires those agents to invoke Decapod throughout governed orientation, task custody, workspace use, validation, proof, and completion. Decapod is therefore a continuing control-plane dependency after one-time initialization, not a command the human repeatedly enters. This contract does not imply interception of every runtime action, and repository markers alone do not prove compliance or outcome improvement.

Decapod has also aided the development of this paper and its executable research artifact since their inception. The repository's governed specifications, tasks, workspaces, validation surfaces, and proof hooks make that use inspectable. This is reflexive dogfooding of the system under study, useful for implementation feedback but not independent validation or a controlled result.

\subsection{Proof and Evidence Boundaries}
Proof definitions contain commands, arguments, descriptions, and required status. A proof run records exit status, duration, pass/fail state, bounded output, run identifiers, actor, and related requirements in the proof event journal. Workunit and workspace-publish gates can require proof results and validation evidence before promotion.

The current implementation therefore supports auditable, machine-checkable evidence, hashes and attestations in selected governance surfaces, and deterministic replay or drift checks where specified. It does not support the stronger certificate described in the earlier draft: this paper does not claim per-command cryptographic diffs, a universal signed \texttt{proof.json}, GPG/SSH signing, arbitrary credential filtering, or complete interception of an agent's PTY. Those are possible future extensions and are excluded from the present empirical claims.
