\section{Evaluation Protocol and Research Program}
\label{sec:evaluation}

The artifact is in harness validation, not empirical data collection. Checked-in records are explicitly \texttt{synthetic\_fixture} data and cannot support a treatment-effect claim. The study's primary test is one-shot walk-away delegation: whether one sufficiently complex natural outcome-oriented request can produce an independently acceptable result without ongoing human coaching because Decapod resolves authoritative project context before implementation inference.

\subsection{Primary Comparison and Mechanism}

The primary comparison uses the same natural delegation prompt, base model and version, decoding and runtime configuration, repository and starting commit, tools, timeout, hidden acceptance rubric, evaluator, and clean-context reset. The conventional cell (CN) receives ordinary repository and tool access without Decapod-mediated constitution resolution, governed orientation, task claim, context capsule, workspace custody, or proof corridor. The Decapod cell (DN) receives the applicable constitution, architecture and specification nodes, capabilities and constraints, standards, prior decisions, task ownership, custody, validation and proof expectations, and escalation rules through the configured corridor. No corrective human coaching is allowed after either run starts; genuine clarification follows only a frozen oracle.

The underlying model is unchanged. Decapod is hypothesized to make the agent system operationally more knowledgeable and contextually aligned by resolving relevant, authoritative project context before implementation inference. The primary interaction is therefore ``type the outcome, start the agent, and walk away,'' not a comparison of prompt length alone.

\subsection{Primary Outcome and Convergence Measures}

The primary outcome is independent fidelity of the final repository state to the human or team's complete intent under zero or tightly bounded post-prompt intervention. A hidden evaluator scores functional correctness, explicit requirements, implicit project invariants, architectural alignment, project conventions, applicable industry standards, regression avoidance, scope discipline, validation completeness, evidence completeness, and readiness for human acceptance.

Convergence efficiency is measured using defined quantities: model inference requests; input and output tokens; tool calls; repository searches; files opened before the first implementation edit; time to the first relevant implementation action; failed implementation attempts; repeated or redundant calls; compiler, test, lint, and validation failure cycles; reversals or substantial rewrites; clarification requests; unresolved questions; elapsed time to acceptable completion; premature completion claims; and cost where available. The paper does not use ``inference hits'' as a measured quantity.

\subsection{Pre-Inference Context Alignment}

For each DN run, the capture path records selected constitutional nodes, project specifications, capability overlays, constraints and negative requirements, standards, prior decisions, unresolved questions, excluded context, total context size, context-resolution latency, and item-level provenance. Each item has an authority source, selection rationale, content hash, size, and selected/excluded status. An independent audit classifies each supplied item as necessary, useful, irrelevant, redundant, contradictory, or harmful. More context is not presumed to be better. CN receives an explicit empty Decapod-context manifest and does not receive the treatment capsule or hidden rubric.

\subsection{Hypotheses and Secondary Ablation}

H1 predicts higher independent intent fidelity for DN than CN. H2 predicts fewer model requests, failed cycles, redundant operations, and substantial rewrites. H3 predicts more application of relevant architecture, conventions, and standards without prompt enumeration. H4 predicts more completion without clarification or corrective intervention. H5 predicts more complete validation, provenance, and review evidence. H6 requires all gains to be evaluated against context-resolution, workspace, validation, token, latency, and operational overhead.

Natural-versus-procedural prompting remains a secondary 2$\times$2 ablation crossing conventional and Decapod-governed substrates with natural and semantically matched procedural prompts. The procedural variant additionally specifies anticipated files, steps, commands, tests, sequencing, and defensive instructions without changing the canonical intent or hidden rubric. A checklist-only condition is secondary only.

\subsection{Tasks, Stages, and Falsification}

Tasks require interacting reasoning: architecture discovery, multi-file or multi-layer changes, compatibility, project conventions, an applicable security, persistence, API, concurrency, or operational standard, validation, and scope discipline. Each package contains canonical intent, motivation, constraints, open questions, escalation policy, a hidden rubric, independent evaluator, paired prompts, clarification oracle, starting commit, available context, interruption/concurrency configuration, and exclusion rules. Runs start from clean repository state and isolated model context.

The program proceeds through harness validation, a labeled pilot, a tagged protocol freeze, confirmatory execution, and independent audit. Pilot records are excluded from confirmatory estimates. Every run identifies its data kind and records model/runtime, repository and Decapod commits, environment, permissions, prompts, events, interventions, oracle responses, context manifest, evaluator result, exclusions, and deviations. Synthetic and pilot records fail closed in confirmatory aggregation.

The delegation thesis is weakened if DN does not improve fidelity, does not reduce failed or redundant cycles, increases cost without improving results, supplies irrelevant or harmful context, still requires regular human correction, is matched by CN at equal or lower effort, benefits from hidden rubric leakage, improves proof while task quality does not improve, or merely relocates supervision into setup. The analysis reports paired task-level contrasts, uncertainty, effect sizes, and task-clustered results; non-inferiority is prohibited until a margin and power analysis are frozen.

\subsection{Study B: Cross-Workbench Handoff Continuity}

Study B terminates Agent A at a frozen observable checkpoint and starts Agent B with a clean context. The credible conventional condition supplies the repository/branch, git history, public task description, and a standardized structured handoff document. The Decapod condition receives the same artifacts plus the implemented approved ownership transfer and frozen Decapod task, context, trajectory, unresolved-state, workspace, and proof references. The first level holds the model fixed where possible; heterogeneous models are an external-validity extension.

RQ7 asks whether governed handoff reduces reorientation and reconstruction; RQ10 asks whether a receiving workbench can continue without transcript replay or an ad hoc human briefing; RQ11 asks whether project-owned state preserves constraints and unresolved questions. H7 predicts fewer reorientation requests and duplicate reads, searches, commands, tests, and decisions. H9 predicts less constraint loss, and H11 predicts less repeated verification when proof remains state-valid. A conventional handoff that performs equally well falsifies the stronger claim.

\subsection{Study C: Concurrent Heterogeneous Fleet}

Study C asks whether multiple independently launched agents can concurrently solve similar but distinct problems in the same codebase while preserving shared authority, distinct custody, isolated mutation, dependency awareness, and integrated proof. It supplies a frozen graph of independent, dependent, duplicate, and similar-but-distinct tasks to at least two agent processes. Similar-but-distinct pairs share architecture, modules, tests, standards, or integration boundaries without expressing the same outcome. Both conditions use separate branches or worktrees. The conventional condition receives normal issue/task descriptions; the treatment additionally uses implemented Decapod claims, labels/categories, dependencies, workspaces, context, events, proof, and publication paths.

RQ8 asks whether custody prevents duplicate starts, RQ9 separates direct collision avoidance from deferred integration failure, and RQ12 evaluates net coordination cost. An independent pre-run audit labels pairwise task relationships and expected overlap. Measures include duplicate starts, false serialization, missed semantic overlap, claim conflicts, redundant discovery, overlapping paths, dependency violations, direct collisions, textual merge conflicts, semantic integration failures, abandoned work, repair effort, duplicated model calls/tokens, wall-clock time, integrated fidelity, proof completeness, and publication success. H8 predicts fewer duplicate starts. H10 predicts fewer direct collisions but explicitly permits no reduction in semantic merge conflicts. Serializing every similar task does not support the concurrency thesis.

\subsection{Study D: Tool-Switch Continuity}

Study D applies Study B's checkpoint and measures to frozen workbench pairs. It records source/destination workbench, model, session, exact human briefing, and the Decapod information delivered to the receiver. Same-model and heterogeneous-model transfers are stratified rather than silently pooled.

\subsection{Study E: Longitudinal Experience Report}

Study E is observational. A checked-in extractor pinned to Decapod \texttt{e7df80d} derives 2,127 commits, 326 tags, 612 merge commits, two explicit revert-subject commits, the February--July 2026 development interval, authorship counts, and changed top-level path frequencies from git. These facts establish sustained repository activity, not a treatment effect, correctness, low review burden, or proof that named workbenches produced particular commits. Author-reported workbench use must be labeled separately. The experience report motivates controlled studies but cannot enter their estimates.

\subsection{Study-Specific Provenance and Authorization}

Run schema version 4 declares one of \texttt{walk\_away\_alignment}, \texttt{handoff\_continuity}, \texttt{concurrent\_fleet}, or \texttt{tool\_switch\_continuity}; Study E uses a separate observational schema. Capture records source/destination model, workbench and session; owner transfer; handoff payload; context before/after; preserved/lost questions, dependencies, and proof; duplicate operations; workspace and claim state; integration and publication outcomes; human briefing; cost; and deviations. Aggregation requires one study, one run kind, and one evidence class. Synthetic, pilot, empirical, and observational records fail closed across those boundaries.

No fleet pilot or empirical run is authorized by this paper revision. Execution requires separately approved task graphs, checkpoints, model/runtime access, credentials, and budget.
