\section{Discussion}
\label{sec:discussion}

The proposed evaluation is designed to test whether a human can delegate one complex outcome and walk away while the agent resolves authoritative context before implementation inference. This is a system-level capability hypothesis, not a claim that Decapod changes the underlying model's parameters or intrinsic intelligence. The primary estimand is independent final-state fidelity and convergence under identical natural prompts; prompt style is secondary.

\subsection{Systems Framing vs. Prompt Engineering}
Historically, the primary method for improving coding agent performance has been prompt engineering—tuning system instructions, adding few-shot examples, or structuring chat outputs. While these methods improve the probability of a model generating correct syntax, they do not change the fundamental execution model. A model executing directly on a production workspace remains unbounded and unverified.

Enforcing explicit custody and proof shifts the focus from inputs (prompts) to execution boundaries and evidence. Decapod can require a claimed task to enter an isolated workspace and can block promotion paths when validation or proof requirements are missing. This is narrower than complete sandboxing: the current kernel does not guarantee that every incorrect change is detected, that every credential is inaccessible, or that every agent action is intercepted. The evaluation must therefore measure both safety benefits and governance overhead.

The findings of Gloaguen et al.~\cite{gloaguen2026agentsmd} sharpen this argument. Their results show that repository context can be respected while still failing to improve task success and increasing inference cost. Instruction following is not equivalent to good governance: an agent may faithfully follow a poisoned, redundant, or over-broad context file. Decapod's constitution and context-capsule structure is motivated by this distinction. It aims to make authority, scope, and proof-relevant context explicit and queryable before inference, while preserving minimality as a design constraint. The independent context audit therefore classifies supplied items as necessary, useful, irrelevant, redundant, contradictory, or harmful. This is a testable architectural hypothesis, not a result established by the present pre-results artifact.

Decapod's current authority chain is concrete: a baseline constitution is embedded in the binary, a project override is explicitly binding, and task context is a selected projection with provenance. The author coined \emph{Intent-Driven Design} as the name of this software-engineering approach in public LinkedIn posts beginning in August 2025 and, by September 2025, was using a constitution to scaffold project boilerplate and derived agent guidance. This design lineage preceded and arose independently of the February 12, 2026 posting of the Gloaguen et al. preprint. The historical attribution does not establish mitigation efficacy: whether structured constitutional authority selects better context or improves outcomes remains untested.

\subsection{The Control Plane for Machine Work}
In traditional software engineering, human developers follow processes: they checkout branches, write tests, request pull request reviews, and run CI/CD pipelines. However, these processes were designed for human timelines. An autonomous agent can run dozens of commands per minute, coordinate across multiple repositories, and execute complex workflows concurrently.

Decapod acts as a local control plane for machine work. It brings intent resolution, task ownership, workspace isolation, validation, proof events, and publication gates into a repository-native workflow. It does not itself coordinate every command issued by an arbitrary model runtime, so the boundary between Decapod evidence and the agent host remains part of the threat model.

For one agent, the central mechanism is pre-inference contextual alignment. For a fleet, the mechanism is reuse: a later process can query durable authority, ownership, selected context, trajectory, and proof instead of treating the previous conversation as the project. This does not require agents to share one orchestrator process. It also creates new risks: stale project memory can spread, claims can become abandoned, and locally valid proof can be reused after its state binding has changed.

\subsection{Cost of Fragmentation}
The economic hypothesis separates nine costs: \emph{context rehydration} (repeated project discovery), \emph{duplicated inference} (multiple agents solving the same work), \emph{collision} (repair after overlapping mutations or assumptions), \emph{handoff} (reconstructing missing state), \emph{verification} (rerunning unavailable or untrusted checks), \emph{integration} (repairing incompatible results), \emph{dialect} (human translation into workbench procedures), \emph{context bloat} (irrelevant governance input), and \emph{abandonment} (work that cannot be resumed economically).

Decapod seeks to lower the cost floor by making useful intent, context, ownership, progress, and evidence reusable across inference requests, sessions, models, workbenches, and teams. This is not yet a result. Every study must count setup, context resolution, state maintenance, workspace, proof, integration, token, latency, and human configuration costs. A simple issue tracker, structured handoff, and worktrees may capture most of the benefit.

A standard agent-callable component could reduce dialect waste further: harnesses would bundle or discover the same versioned governance semantics while humans continue delegating in natural language. That is a portability hypothesis, not current deployment. A Decapod-specific CLI may itself become lock-in unless the profile, conformance vectors, adapters, and failure behavior can be implemented independently.

\subsection{Trust, Auditing, and the Limitations of Proof}
A central hypothesis of this study is that structured proof files and trajectories will reduce human review time. In a transcript-only workflow, human reviews are cognitively demanding: the developer must read long chat threads, try to understand why the agent chose a specific path, and then manually check if the code works. A Decapod run, however, separates the conversational noise from the systems evidence.

Programmatic proof has strict limits. Decapod evidence can show that configured commands ran and what they returned, but it cannot verify soft requirements such as code elegance, architectural maintainability, or completeness of an unstated user goal. The goal is therefore to make objective checks and governance state easier to inspect, not to replace human review or convert a validation pass into a security guarantee. The thesis is weakened if natural delegation performs no better, if procedural prompts remain necessary, if governance relocates human effort into setup, or if overhead outweighs reliability and autonomy benefits.

Fleet coherence has equally strict limits. Exclusive claims can reduce duplicate starts while doing nothing for task decomposition quality. Worktrees can eliminate direct file trampling while deferring semantic conflicts to integration. Handoff records can improve attribution while omitting the decision the receiver needs. Proof can become more portable while actual task quality stays unchanged. Null and negative results on these dimensions would narrow the systems thesis even if the mechanisms operate as implemented.
