\section{Related Work}
\label{sec:relatedwork}

Our research intersects with coding-agent workbenches, multi-agent orchestration, repository context and memory, task/workspace coordination, validation, and software-supply-chain evidence.

\subsection{LLM Agents in Software Engineering}
Since scalable Transformers~\cite{vaswani2017attention} and few-shot models such as GPT-3~\cite{brown2020language}, code generation has moved from autocompletion to read-write-execute agents. OpenHands provides a model-flexible software-agent platform with command-line, browsing, sandbox, benchmark, and multi-agent coordination surfaces~\cite{wang2024openhands}. These systems overlap substantially with Decapod on realistic execution and isolation.

Our narrower focus is the project control plane external to one agent loop. The research question is whether repository-native authority, ownership, handoff, context, and proof remain usable when independently launched processes do not share one conversation or orchestrator lifecycle. We do not claim that all contemporary agents are transcript-only or lack memory, sandboxing, or coordination.

\subsection{Multi-Agent Orchestration and Role-Based Teams}
MetaGPT encodes software-development roles and standard operating procedures into a multi-agent framework~\cite{hong2023metagpt}. ChatDev organizes specialized agents into a chat chain across design, coding, testing, and documentation~\cite{qian2024chatdev}. AutoGen provides conversable-agent abstractions for composing multi-agent applications and human/tool interaction~\cite{wu2023autogen}. OpenHands also supports coordination among agents inside its platform~\cite{wang2024openhands}.

These frameworks show that role specialization, conversation topology, model flexibility, memory, and sandboxed execution are not unique to Decapod. The proposed distinction is architectural rather than categorical: Decapod externalizes governance into durable project state and a CLI/RPC contract so separately launched workbenches can participate without joining one chat chain or long-running orchestrator process. Whether that externalization improves handoff or concurrency is precisely what Studies B--D must test. A structured baseline may match it.

\subsection{Repository Context, Context Poisoning, and Memory}
Gloaguen et al.'s study of repository-level agent instructions is the closest contemporaneous empirical comparison~\cite{gloaguen2026agentsmd}. Across multiple agents and LLMs, that study reports no improvement in task success, more than 20\% higher inference cost, broader exploration such as additional testing and file traversal, and substantial instruction-following. Its result is important for this paper's thesis: repository context is not inert documentation. It changes the agent's execution trajectory and resource consumption, so unnecessary or contradictory requirements can act as context poisoning even when the agent appears to follow them.

Our position is not that a longer constitution is automatically better. Decapod responds to the same risk by treating context as governed control-plane input: the constitution is structured into queryable nodes, \texttt{context.resolve} and context capsules scope what is supplied before inference, and plans preserve unknowns, human questions, stop conditions, and proof expectations. This design makes context selection and authority inspectable rather than placing an unbounded instruction file in the prompt. The ETH Zürich findings therefore support the need to evaluate context quality, scope, cost, and behavioral side effects; they do not by themselves establish that Decapod's structure improves task success. Our evaluation should test whether structured, scoped context reduces harmful drift and review burden while avoiding the overhead identified by Gloaguen et al.

\subsection{Program Verification and Sandboxing}
Our custody and proof primitives build on program verification and systems sandboxing. Traditional verification systems target mathematical correctness; our proof primitive is pragmatic, configured evidence from compilers, test suites, static analysis, workunits, and provenance.

Sandboxing platforms isolate untrusted execution, while Git worktrees support multiple branches in separate working directories~\cite{gitworktree}. Decapod composes task claims, worktree checks, and optional containers. Worktrees isolate mutations but do not isolate credentials or prove merge compatibility; this paper does not call them a shell sandbox.

\subsection{Workflow Durability, Task Coordination, and Provenance}
Issue trackers and workflow engines preserve task and execution state independently of one worker. Software supply-chain systems such as in-toto go further by binding authorized steps, actors, materials, and products into verifiable metadata~\cite{torresarias2019intoto}. Decapod's proof and provenance surfaces occupy a narrower development-time role: they record configured governance/validation evidence for agent work. They are not a substitute for in-toto's cryptographic supply-chain guarantees, nor do they sign every action.

The closest conventional controls for the fleet studies are therefore not ungoverned shared directories. They are issue/task descriptions, structured handoff documents, separate worktrees, git history, CI, and provenance tooling. The experiments compare Decapod against these credible combinations.

\subsection{Agent and Protocol Interoperability}
MCP specifies a host--client--server architecture, JSON-RPC data layer, capability negotiation, tools/resources/prompts, and standard local or HTTP transports~\cite{mcp2025architecture}. A2A specifies discovery and task/artifact exchange between independent, opaque agent systems~\cite{a2a2026spec}. They address adjacent interoperability layers: MCP connects an agent host to callable context/tools, while A2A connects collaborating agents. HTTP and IP demonstrate the broader value of narrow interfaces that permit intermediaries, layered transports, and independent implementations~\cite{rfc9110http,rfc791ip}.

Decapod does not currently implement these protocols. Its present interface is a local CLI and Decapod-specific structured process RPC. The proposed contribution is a candidate governance profile over constitutional authority, task custody, trajectory, proof, and publication, with MCP and A2A adapters rather than a competing all-purpose agent protocol. The claim must remain conditional until an independent implementation, conformance suite, identity model, and actual harness integrations exist.

\subsection{Distributed Systems and Ledger Audits}
The framing of agentic coding as a distributed systems execution problem is inspired by classic distributed consensus and audit architectures. Lamport's happenings-before relation~\cite{lamport1978time} laid the groundwork for ordering events across independent computing nodes. MapReduce~\cite{dean2004mapreduce} and distributed databases demonstrated how to partition state and manage concurrency across large scale clusters.

Similarly, event sourcing and cryptographically bound artifacts provide conceptual inspiration for Trajectory and Proof. Decapod's current contribution is narrower: it journals selected governance/proof events and exposes validation, provenance, and attestation surfaces. It does not implement distributed consensus, log every tool call, or sign one universal execution ledger.
