\section{Conclusion}
\label{sec:conclusion}

As LLM-based coding agents transition from suggestion to delegated execution, we must treat agentic software engineering as a distributed systems execution problem rather than a simple text-generation task. The traditional transcript-only workflow—relying solely on chat transcripts and code diffs—fails to provide the safety, recovery, and verification guarantees required for enterprise codebases.

This paper formalized four primitives for accountable agentic execution:
\begin{itemize}
    \item \textbf{Intent}: Human/team desired outcomes, motivations, constraints, priorities, unknowns, stop conditions, and completion standards, progressively resolved through governed artifacts.
    \item \textbf{Custody}: Bounded execution sandboxes and permission limits.
    \item \textbf{Trajectory}: Event-backed governance records rendered for reconstruction and review.
    \item \textbf{Proof}: Machine-checkable validation, proof-run, workunit, and provenance evidence.
\end{itemize}

We presented Decapod, a daemonless, local-first governance kernel, as a reference implementation pinned to commit \texttt{207e9a3}. The implemented binary enables pre-inference context alignment, allowing a human to delegate complex desired outcomes and walk away with less procedural cognitive load. The primary research program tests whether that mechanism measurably preserves intent fidelity and reduces correction relative to credible controls. The expanded program tests fleet coherence: whether heterogeneous, independently launched processes can preserve common authority, own separate work, hand tasks across workbenches, solve similar but distinct problems concurrently, and share state-bound evidence without depending on one conversation. The constitution---embedded baseline plus project override in the current implementation---is the authority root from which task context is selected.

We additionally offer these governance semantics as a candidate agent-callable interoperability profile that coding-agent harnesses could pre-bundle or discover. The goal is for humans to delegate naturally while agents invoke a portable machine contract, not for users to learn another vendor dialect. Native MCP/A2A adapters, HTTP transport, organization policy, conformance, and actual vendor bundling remain future work.

The comparative-effect claims are prospective. Controlled records remain synthetic; the longitudinal git snapshot is observational and non-causal. Decapod demonstrates real local claims, handoff, context, memory, workspace, event, proof, and publication mechanisms, but not complete security, lossless conversation transfer, conflict-free integration, global consensus, or production cloud federation. The research program is designed to reject the thesis if credible structured handoffs and ordinary worktrees perform equally well, shared context harms agents, or coordination overhead outweighs measured benefit. The author has not observed those failure conditions while using Decapod to build Decapod since early February 2026 or across subsequent repositories initialized once and then continuously governed through their agent contracts; that experience is observational and author-reported rather than a controlled comparative result.
