\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[margin=1.2in]{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{microtype} % Improves typographical justification and spacing
\usepackage{cite}
\usepackage{url}
\usepackage{xcolor}
\usepackage{hyperref}

% Typography (Palatino for body text, Helvetica for sans-serif headers)
\usepackage{mathpazo}
\usepackage[scaled=0.92]{helvet}
\usepackage{courier}

% Customize section headings
\usepackage{sectsty}
\allsectionsfont{\sffamily\bfseries\color[HTML]{1e293b}} % Slate-gray headings

\definecolor{indigo-slate}{HTML}{1e1b4b} % Deep Indigo Slate
\hypersetup{
    colorlinks=true,
    linkcolor=indigo-slate,
    citecolor=indigo-slate,
    filecolor=indigo-slate,
    urlcolor=indigo-slate,
    pdftitle={Intent, Custody, Trajectory, and Proof: Toward Accountable Execution in Agentic Software Engineering},
    pdfauthor={Alex H. Raber (Decapod Labs)},
}

\title{\Huge\sffamily\bfseries Intent, Custody, Trajectory, and Proof:\\\Large Toward Accountable Execution in Agentic Software Engineering}

\author{\textbf{Alex H. Raber}\\[0.5em]
\sffamily Decapod Labs\\[0.2em]
\texttt{alexhraber@gmail.com}
}
\date{\sffamily Technical Report --- June 2026}

\begin{document}

\maketitle

\begin{abstract}
\noindent Large language model (LLM) agents are transitioning from conversational programming assistants to autonomous executors capable of editing codebases, running build tools, executing tests, and managing multi-file software deployments. Despite this shift, current agent workflows still rely on standard prompt transcripts and code diffs as the primary logs of execution. As agent runs become concurrent, long-running, and tool-intensive, this transcript-only paradigm exposes repositories to significant failure modes: lost intent, authority leakage, concurrent state corruption, and unverifiable completion claims.

\medskip
\noindent This paper frames agentic software engineering as a distributed execution problem. We introduce a framework built around four primitives to achieve accountable execution: \emph{Intent} (durable target objectives), \emph{Custody} (isolated environment bounds and tool permissions), \emph{Trajectory} (immutable, kernel-logged audit trails), and \emph{Proof} (cryptographically anchored, machine-checkable evidence of completion). We present Decapod, a local-first governance kernel, as a reference implementation of this model. We propose an empirical evaluation protocol to measure governed execution across 35 multi-step software engineering tasks against prompt-only and checklist-disciplined baselines. We outline how explicit custody and verification gates are hypothesized to reduce false completion claims, prevent concurrent edit conflicts, and decrease the human developer effort required to audit and merge autonomous code contributions.
\end{abstract}

\vspace{1.5em}

% Inputting sections
\input{sections/01-introduction}
\input{sections/02-background}
\input{sections/03-problem}
\input{sections/04-model}
\input{sections/05-system}
\input{sections/06-evaluation}
\input{sections/07-discussion}
\input{sections/08-related-work}
\input{sections/09-conclusion}

\bibliographystyle{abbrv}
\bibliography{references}

\end{document}
